---
title: "Lab 7"
author: "Mark Tello-Rincon"
format: html
editor: visual
embed-resources: true
theme: cerulean
---

#### Set-up

```{r}
library(httr)
library(xml2)
library(stringr)
```

### Question 1

How many Sars-Cov-2 papers?

```{r}
## website
website <- xml2::read_html("https://pubmed.ncbi.nlm.nih.gov/?term=sars-cov-2")

## counts
counts <- xml2::xml_find_first(website, "//span[@class='value']")

## turn into text
counts <- as.character(counts)

## extracting data using regex
totalcount <- str_extract(counts, "[0-9,]+")

## remove commas to convert to numeric
totalcount <- gsub(",", "", totalcount)
totalcount <- as.numeric(totalcount)
print(totalcount)
```

### Question 2

```{r}
## text file
abstracts <- readLines('~/Library/CloudStorage/GoogleDrive-tellorin@usc.edu/.shortcut-targets-by-id/10yI1Vp2x44iBX7T-_NfWNeL7go8kwnUH/2. College - USC/1. Degree/1. Courses/Y4 Senior/Fall 2025/PM 566/PM566-Repository/Lab/Lab 7/Lab 7/data/abstract-sars-cov-2-set.txt', warn = FALSE)

## combine all text into one character
abstracts <- paste(abstracts, collapse = '\n')

## split text whenever 3 new lines occur in a row (two blank lines)
abstracts <- unlist(strsplit(abstracts, split = '\n\n\n'))

## replace any remaining "\n" symbols with spaces
abstracts <- gsub("\n", " ", abstracts)

## replace multiple spaces with single space
abstracts <- gsub(" +", " ", abstracts)
```

### Question 3

What are the 10 most common institutions? Discuss how you could you improve these results.

-   The ten most common institutions are the University of California, University of Pennsylvania, and Stanford University, but also contain noisy entries like `from University` and `at University`. To improve the results, we could refine the regex to use full, capitalized institution patterns, standardize variations in names like `UC ____` and `University of California, ____`, remove trailing punctuation or prepositions, and grouping similar names.

```{r}
institution <- str_extract_all(
  abstracts,
  "([[:alpha:]-]+\\s+University|University\\s+of\\s+[[:alpha:]-]+|[[:alpha:]-]+\\s+Institute\\s+of\\s+[[:alpha:]-]+)"
) 
institution <- unlist(institution)

head(sort(table(institution), decreasing = TRUE), 10)
```

### Question 4

```{r}
abstracts <- readLines('~/Library/CloudStorage/GoogleDrive-tellorin@usc.edu/.shortcut-targets-by-id/10yI1Vp2x44iBX7T-_NfWNeL7go8kwnUH/2. College - USC/1. Degree/1. Courses/Y4 Senior/Fall 2025/PM 566/PM566-Repository/Lab/Lab 7/Lab 7/data/abstract-sars-cov-2-set.txt', warn = FALSE)
abstracts <- paste(abstracts, collapse = '\n')
abstracts <- unlist(strsplit(abstracts, split = '\n\n\n'))

journal <- str_extract(abstracts, "^[0-9]+\\.\\s[^\n]+")

titles <- sapply(abstracts, function(x){
  unlist(strsplit(x, split = "\n\n"))[2]
}, USE.NAMES = FALSE)

authors <- sapply(abstracts, function(x){
  unlist(strsplit(x, split = "\n\n"))[3]
}, USE.NAMES = FALSE)

affiliations <- str_extract(abstracts, "Author information:[\\s\\S]*?(?=\\n\\n|$)")

papers <- data.frame(
  Journal = journal,
  Title = titles,
  Authors = authors,
  Affiliations = affiliations,
  stringsAsFactors = FALSE
)

knitr::kable(papers[1:5, ])
```
