---
title: "Homework 4"
author: "Mark Tello-Rincon"
format: html
editor: visual
embed-resources: true
theme: darkly
---

##### HPC

```{r}
# Total for each row
fun1 <- function(mat) {
  n <- nrow(mat)
  ans <- double(n) 
  for (i in 1:n) {
    ans[i] <- sum(mat[i, ])
  }
  ans
}

fun1alt <- function(mat) {
  rowSums(mat)
}

# running cumulative sum by row
fun2 <- function(mat) {
  n <- nrow(mat)
  k <- ncol(mat)
  ans <- mat
  for (i in 1:n) {
    for (j in 2:k) {
      ans[i,j] <- mat[i, j] + ans[i, j - 1]
    }
  }
  ans
}

fun2alt <- function(mat) {
  t(apply(mat, 1, cumsum))
}
```

###  Question 1

-   The optimized functions produced identical outputs to the originals
-   The `rowSums()` version of `fun1` ran about 30 times faster than the loop
-   The `apply(cumsum)` version of `fun2` was about 5 times faster
-   Vectorization significantly improves performance/speed without changing the results

```{r, warning=FALSE}
# Use the data with this code
set.seed(2315)
dat <- matrix(rnorm(200 * 100), nrow = 200)

# Test for the first
microbenchmark::microbenchmark(
  fun1(dat),
  fun1alt(dat), unit = "relative"
)

# Test for the second
microbenchmark::microbenchmark(
  fun2(dat),
  fun2alt(dat), unit = "relative"
)

# Estimate pi thru sim
sim_pi <- function(n = 1000, i = NULL) {
  p <- matrix(runif(n*2), ncol = 2)
  mean(rowSums(p^2) < 1) * 4
}

# Example of the run
set.seed(156)
sim_pi(1000) # 3.132

# This runs the simulation a 4,000 times, each with 10,000 points
set.seed(1231)
system.time({
  ans <- unlist(lapply(1:4000, sim_pi, n = 10000))
  print(mean(ans))
})
```

###  Question 2

-   `parLapply()` produced roughly the same estimate of pi and was 2 times faster

```{r, echo = FALSE}
library(parallel)
```

```{r}
cl <- makeCluster(4)

clusterExport(cl, varlist = c("sim_pi"), envir = environment())

clusterSetRNGStream(cl, 1231)

system.time({
  ans <- unlist(parLapply(cl, 1:4000, function(i) sim_pi(n = 10000)))
  print(mean(ans))
})

stopCluster(cl)
```

##### SQL

```{r}
library(RSQLite)
library(DBI)

# Initialize a temporary in memory database
con <- dbConnect(SQLite(), ":memory:")

# Download tables
film <- read.csv("https://raw.githubusercontent.com/ivanceras/sakila/master/csv-sakila-db/film.csv")
film_category <- read.csv("https://raw.githubusercontent.com/ivanceras/sakila/master/csv-sakila-db/film_category.csv")
category <- read.csv("https://raw.githubusercontent.com/ivanceras/sakila/master/csv-sakila-db/category.csv")

# Copy data.frames to database
dbWriteTable(con, "film", film)
dbWriteTable(con, "film_category", film_category)
dbWriteTable(con, "category", category)
```

###  Question 3

-   There are 5 rating categories in the dataset
-   PG‑13 has the largest number of movies with 223
-   Rated R movies had the least with 195

```{sql connection=con}
SELECT rating, COUNT(*) AS num_movies
FROM film
GROUP BY rating;
```

###  Question 4

-   PG films have the lowest average replacement cost at about 18.96
-   PG‑13 films have the highest at about 20.40
-   Rental rates are all close to 3, with PG films having the highest average rental rate at 3.05

```{sql connection=con}
SELECT rating,
       AVG(replacement_cost) AS avg_replacement_cost,
       AVG(rental_rate) AS avg_rental_rate
FROM film
GROUP BY rating;
```

###  Question 5

-   Category 9 has the most among the first ten shown with 73 films
-   Category 4 has the least with 57 films

```{sql connection=con}
SELECT fc.category_id, COUNT(*) AS num_films
FROM film_category fc
INNER JOIN film f ON f.film_id = fc.film_id
GROUP BY fc.category_id;
```

###  Question 6

-   `Sports` is the most popular category with 74 films

```{sql connection=con}
SELECT c.name, COUNT(*) AS num_films
FROM film_category fc
INNER JOIN film f ON f.film_id = fc.film_id
INNER JOIN category c ON c.category_id = fc.category_id
GROUP BY c.name
ORDER BY num_films DESC
LIMIT 1;
```
